Late and early effective start of non-pharmaceutical interventions inferred during COVID-19 outbreaks

### 2020-05-11 ms/ms.tex
revised median, highlights for todos

The likelihood function is defined in \autoref{eq:likelihood}.
We define the following prior distributions on the model parameters $P(\theta)$: 
\begin{equation} \label{eq:priors}
\begin{aligned} % TODO verify before submitting
Z & \sim \mathit{Uniform}(2, 5) \\
D & \sim \mathit{Uniform}(2, 5) \\
\mu & \sim \mathit{Uniform}(0.2, 1) \\
\beta & \sim \mathit{Uniform}(0.8, 1.5) \\
\lambda & \sim \mathit{Uniform}(0, 1) \\
\alpha_1, \alpha_2 & \sim \mathit{Uniform}(0.02, 1)\\
E(0) & \sim \mathit{Uniform}(0, 3000) \\
I_u(0) & \sim \mathit{Uniform}(0, 3000) \\
\tau &\sim \mathit{TruncatedNormal}(\tau^*, 5, 1, T-2),
\end{aligned}
\end{equation}
where $\mathit{TruncatedNormal}(\mu, \sigma, a, b)$ is a truncated normal distribution with mean $\mu$ and standard deviation $\sigma$ taking values between $a$ and $b$; $T$ is the number of days in the data $\vec{X}$; and $\tau^*$ is the official start of the NPI.
Most priors follow \citet{Li2020}, with the following exceptions.
$\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood and unreasonable parameter estimates.
For the effective start of NPIs $\tau$ we have also tested an uninformative uniform prior $U(1,T-1)$. DIC (see definition below) was lower for the truncated normal prior in most countries, except \hl{Germany?}. More importantly, the uninformative prior could result in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom (this was due to MCMC chains being stuck in low posterior regions of the parameter space).
We therefore decided to use the more informative truncated normal prior.

@@

Several studies have described the effects of non-pharmaceutical interventions in different geographical regions~\citep{Flaxman2020,Gatto2020,Li2020}. 
These studies have assumed that the parameters of the epidemiological model change at a specific date, as in \autoref{eq:NPI_model}, and set the change date $\tau$ to the official NPI date $\tau^*$ (\autoref{table:table1}).
They then fit the model once for time $t<\tau^*$ and once for time $t \ge \tau^*$.
For example, \citet{Li2020} estimate the dynamics in China before and after $\tau^*$ at Jan 23. Thereby, they effectively estimate $(\beta, \alpha_1)$ and $(\lambda, \alpha_2)$ separately.
Here we estimate the posterior distribution $P(\tau \mid \vec{X})$ of the \emph{effective} start date of the NPIs by jointly estimating $\tau, \beta, \lambda, \alpha_1, \alpha_2$ on the entire data per region (e.g. Italy, Austria), rather than splitting the data at $\tau^*$.
We then estimate the maximum a posteriori (MAP) estimate $\hat{\tau}=\mathit{argmax}_{\tau}{P(\tau \mid \vec{X})}$. \hl{using median or argmax?}

We find that a model that considers an NPI (\autoref{eq:NPI_model}) is a better fit to the data than a model without an NPI, i.e. with constant $\beta$ and $\alpha$ (\hl{$\Delta DIC > ?$ for all regions}.)
We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in most regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}): the \hl{95\%} credible interval on $\hat{\tau}$ does not include $\tau^*$, and the DIC of the model with free $\tau$ parameter is lower than that of a model with a fixed $\tau \equiv \tau^*$ (\hl{$\Delta DIC > ?$}.) The exception that proves the rule is \hl{Switzerland}, where the effective and official dates are the same. Another important exception is the \hl{United~Kingdom?} % TODO need more data? Maybe not such an exception?

@@

In Italy, the first case officially confirmed on Feb~21, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
That is, the official date $\tau^*$ is either Mar~8 or~11.
However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.7 days 95\% CI ; \autoref{fig:fig2}).
Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate  the effective start of NPIs to be several days layer at around Mar~2 ($\pm$2.65 days 95\% CI \autoref{fig:fig2}).

%%% Fig 2 %%%
\begin{figure}[h]
    \centering
    % Wuhan
    % τ_hat = 19.33
	% +-2.65 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig2a.pdf}
    \end{subfigure}
  	~
	% Italy
	% τ_hat = 24.74
	% +-0.71 days for 0.95 quantile
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig2b.pdf}
    \end{subfigure}
    \caption{
	\textbf{Late effect of non-pharmaceutical interventions in Italy and Wuhan, China.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
    }
    \label{fig:fig2}
\end{figure}

%%% Early
\paragraph*{Early effective start of NPIs.}
In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:fig1}).
In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (2:1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ at Mar~8 or~9 ($\pm$2.96 95\%CI), rather than Mar~14 (\autoref{fig:fig3}).

Similarly, in France the official lockdown started at Mar~17 ($\tau^*$), with initial NPIs at Mar 13~\citep{Flaxman2020}. 
However, we estimate the effective start of NPIs $\hat{\tau}$ at Mar~8 ($\pm$5.9 days 95\% CI). Although the credible interval is wide, spanning from Mar~2 to Mar~13, the official lockdown start at Mar~17 is later still (\autoref{fig:fig3}).

Interestingly, the effective start of NPIs $\hat{\tau}$ in both France and Spain is estimated at Mar~8, although the official dates are differ by three days. Moreover, the number of daily cases was similar until Mar~8 in both countries, but diverged by Mar~13, reaching significantly higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).

@@

%%% Exact
\paragraph*{The exception that proves the rule.}
We find one case in which the official and effective dates match: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
Indeed, our MAP estimate $\hat{\tau}$ is Mar~20, and the posterior distribution shows two density peaks: a smaller one between Mar~10 and Mar~14, and a taller one between Mar~17 and Mar~22. It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.

>>>>

The likelihood function is defined in \autoref{eq:likelihood}.
We define the following prior distributions on the model parameters $P(\theta)$: 
\begin{equation} \label{eq:priors}
\begin{aligned} % TODO verify before submitting
Z & \sim \mathit{Uniform}(2, 5) \\
D & \sim \mathit{Uniform}(2, 5) \\
\mu & \sim \mathit{Uniform}(0.2, 1) \\
\beta & \sim \mathit{Uniform}(0.8, 1.5) \\
\lambda & \sim \mathit{Uniform}(0, 1) \\
\alpha_1, \alpha_2 & \sim \mathit{Uniform}(0.02, 1)\\
E(0) & \sim \mathit{Uniform}(0, 3000) \\
I_u(0) & \sim \mathit{Uniform}(0, 3000) \\
\tau &\sim \mathit{TruncatedNormal}(\tau^*, 5, 1, T-2),
\end{aligned}
\end{equation}
where $\mathit{TruncatedNormal}(\mu, \sigma, a, b)$ is a truncated normal distribution with mean $\mu$ and standard deviation $\sigma$ taking values between $a$ and $b$; $T$ is the number of days in the data $\vec{X}$; and $\tau^*$ is the official start of the NPI.
Most priors follow \citet{Li2020}, with the following exceptions.
$\lambda$ is used to ensure transmission rates are lower after the start of the NPIs ($\lambda < 1$).
We checked values of $\Delta t$ larger than five days and found they generally produce lower likelihood, higher DIC (see below), and unreasonable parameter estimates.
For the effective start of NPIs $\tau$ we have also tested an uninformative uniform prior $U(1,T-1)$. DIC (see below) was lower for the truncated normal prior in \hl{most countries}. More importantly, the uninformative prior could result in non-negligible posterior probability for unreasonable $\tau$ values, such as Mar~1 in the United~Kingdom. This was probably due to MCMC chains being stuck in low posterior regions of the parameter space.
We therefore decided to use the more informative truncated normal prior.

@@

Several studies have described the effects of non-pharmaceutical interventions in different geographical regions~\citep{Flaxman2020,Gatto2020,Li2020}. 
These studies have assumed that the parameters of the epidemiological model change at a specific date, as in \autoref{eq:NPI_model}, and set the change date $\tau$ to the official NPI date $\tau^*$ (\autoref{table:table1}).
They then fit the model once for time $t<\tau^*$ and once for time $t \ge \tau^*$.
For example, \citet{Li2020} estimate the dynamics in China before and after $\tau^*$ at Jan 23. Thereby, they effectively estimate $(\beta, \alpha_1)$ and $(\lambda, \alpha_2)$ separately.
Here we estimate the posterior distribution $P(\tau \mid \vec{X})$ of the \emph{effective} start date of the NPIs by jointly estimating $\tau, \beta, \lambda, \alpha_1, \alpha_2$ on the entire data per region (e.g. Italy, Austria), rather than splitting the data at $\tau^*$.
We then estimate the posterior probability $P(\tau \mid \vec{X})$ by marginalizing the joint posterior, and estimate $\hat{\tau}$ as the posterior median.

We find that a model that considers an NPI (\autoref{eq:NPI_model}) is a better fit to the data than a model without an NPI, i.e. with constant $\beta$ and $\alpha$ (\hl{$\Delta DIC > ?$ for all regions}.)
We compare the official $\tau^*$ and effective $\hat{\tau}$ start of NPIs and find that in some regions the effective start of NPI  significantly differs from the official date (\autoref{fig:fig1}): the credible interval on $\hat{\tau}$ does not include $\tau^*$, and the DIC of the model with free $\tau$ parameter is lower than that of a model with a fixed $\tau \equiv \tau^*$ (\hl{$\Delta DIC > ?$}.)

@@

\hl{In Italy}, the first case officially confirmed on Feb~21, a lockdown was declared in Northern Italy on Mar~8, with social distancing implemented in the rest of the country, and the lockdown was extended to the entire nation on Mar 11~\citep{Gatto2020}.
That is, the official date $\tau^*$ is either Mar~8 or~11.
However, we estimate the effective date $\hat{\tau}$ at Mar~16 ($\pm$0.7 days 95\% CI ; \autoref{fig:fig2}).
Similarly, \hl{in Wuhan, China}, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate  the effective start of NPIs to be several days layer at around Mar~2 ($\pm$2.65 days 95\% CI \autoref{fig:fig2}).

%%% Fig 2 %%%
\begin{figure}[h]
    \centering
    % Wuhan 
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig2a.pdf}
    \end{subfigure}
  	~
	% Italy
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{../figures/Fig2b.pdf}
    \end{subfigure}
    \caption{
	\textbf{Late effect of non-pharmaceutical interventions in Italy and Wuhan, China.}
    Posterior distribution of $\tau$, the effective start date of NPI, is shown as a histogram of MCMC samples. Red line shows the official last NPI date $\tau^*$. Black line shows the estimated $\hat{\tau}$. Shaded area shows a 95\% credible interval (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$). 
    }
    \label{fig:fig2}
\end{figure}

%%% Early
\paragraph*{Early effective start of NPIs.}
In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:fig1}).
In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (2:1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the \hl{effective start of NPI} $\hat{\tau}$ at Mar~8 or~9 ($\pm$2.96 95\%CI), rather than Mar~14 (\autoref{fig:fig3}).

Similarly, in \hl{France} the official lockdown started at Mar~17 ($\tau^*$), with initial NPIs at Mar 13~\citep{Flaxman2020}. 
However, we estimate the effective start of NPIs $\hat{\tau}$ at Mar~8 ($\pm$5.9 days 95\% CI). Although the credible interval is wide, spanning from Mar~2 to Mar~13, the official lockdown start at Mar~17 is later still (\autoref{fig:fig3}).

Interestingly, the effective start of NPIs $\hat{\tau}$ in both France and Spain is estimated at Mar~8, although the official dates are differ by three days. Moreover, the number of daily cases was similar until Mar~8 in both countries, but diverged by Mar~13, reaching significantly higher numbers in Spain (\autoref{fig:fig-fr-vs-es}).
This may suggest that correlation exist between effective start in NPIs due to global or international events.

@@

%%% Exact
\paragraph*{The exception that proves the rule.}
We find one case in which the official and effective dates match: Switzerland ordered a national lockdown on Mar~20, after banning public evens and closing schools on Mar~13 and~14~\citep{Flaxman2020}.
Indeed, we estimate that $\hat{\tau}$ is \hl{Mar~20}, and the posterior distribution shows two density peaks: a smaller one between Mar~10 and Mar~14, and a taller one between Mar~17 and Mar~22. It's also worth mentioning that Switzerland was the first to mandate self isolation of confirmed cases~\citep{Flaxman2020}.
